---
title: "Load NHANES"
output: html_notebook
---

Glen Dec 2024

Read NHANES fiels

```{r}
gRootName = "nhanes"
gRootDir = "C:/Users/Glen/OneDrive - Dalhousie University/Documents"  #where scripts are
outputDir = "C:/postdoc/nhanes" #analysis directory
dataDir =  "C:/data/nhanes"
#source(sprintf("%s/logreg.R",gRootDir),verbose=0) 
#source(sprintf("%s/pca.R",gRootDir),verbose=0)  #cov2
source(sprintf("%s/main.R",gRootDir),verbose=0)
source(sprintf("%s/plots.R",gRootDir),verbose=0) #TilePlot
source(sprintf("%s/survival.R",gRootDir),verbose=0) #ArrayToStartStop
#source(sprintf("%s/sf.R",gRootDir),verbose=0) 

library(tidyr) #for spread (long to wide) and gather (wide to long) #http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/
library(GGally)
library(gridExtra)
library(ggrepel)
library(survival)
#library(RNHANES) #not working


library(rvest)
library(httr)
library(stringr)
library(haven)
library(dplyr)

```


```{r}
# Base URL for NHANES
nhanes_base_url <- "https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx"

# Function to get all file URLs for a specific cycle
get_file_urls <- function(component="Questionnaire") {
  # Construct the URL for the cycle
  #cycle_url <- sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=%s&CycleBeginYear=%s",component, substr(year, 1, 4))
  #cycle_url <- sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=%s&CycleBeginYear=%s",component, year)
  cycle_url <- sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=%s",component)
  
  
  # Scrape the page to find .XPT file links
  page <- read_html(cycle_url)
  links <- page %>%
    html_nodes("a") %>%
    html_attr("href") %>%
    str_subset("(?i)\\.XPT$") #?i makes it ignore case
  
  #print(cycle_url)
  #print(links)
  if (length(links) == 0) {
      warning(sprintf("No .XPT links found for component %s.", component))
  }
  
  # Create full URLs
  links <- paste0("https://wwwn.cdc.gov", links)
  return(links)
}

# Function to download all lab files for a specific cycle
download_and_save_files <- function( dest_dir,final_dir,removeXPT=T,component="Questionnaire",name="clinic",maxSizeGB=.1) {
  file_urls <- get_file_urls(component=component)
  dir.create(final_dir, showWarnings = FALSE)
  dir.create(dest_dir, showWarnings = FALSE)
  
  
  ldata = list()
  vars = character()
  ids = character()
  ind = 1
  for (url in file_urls) 
  {
    file_name <- basename(url)
    xpt_path <- file.path(dest_dir, file_name)  # Temporary XPT file path
    csv_path <- file.path(dest_dir, paste0(tools::file_path_sans_ext(file_name), ".csv"))


        #check size before downloading
    response = HEAD(url)
    sizeGB = as.numeric(headers(response)$`content-length`)/10^9    

    if(sizeGB > maxSizeGB)
    {
      warning(sprintf("%s was too large (size: %.3f GB, max: %.3f GB), skipping...",xpt_path,sizeGB,maxSizeGB))
      next
    }
    
    message("Downloading: ", file_name)
      

    
    # Download the .XPT file
    GET(url, write_disk(xpt_path, overwrite = TRUE))
    
    # Read the XPT file and save as CSV
    data <- read_xpt(xpt_path)
    data = as.data.frame(data)
    if(!("SEQN"%in%colnames(data)))
    {
      message("SEQN not found, skipping...")
      if(removeXPT) file.remove(xpt_path)
      next #skips instead
      #data[,"SEQN"] = sprintf("%s0%d0%.0f",as.character(data[,1]),1:nrow(data),10^(ind)) #generates SEQN
    }
    #drop duplicates
    logi = duplicated(data[,"SEQN"])
    if(sum(logi)>0)
    {
      warning(sprintf("dropping %d repeated individuals",sum(logi)))
      data = data[!logi,]
    }
    rownames(data)=data[,"SEQN"]
    ldata[[ind]] = data
    vars = c(vars,colnames(data))
    ids = c(ids,data[,"SEQN"])
    ind = ind + 1
      
    #print(table(is.na(data[,"RIDAGEYR"]))) #temp - where is the age info? 0 NAs here
  
      
    # Remove the temporary XPT file
    if(removeXPT) file.remove(xpt_path)
  }
  

  
  
  ids = unique(ids)
  vars = unique(c("SEQN",vars))

  merged_data <- data.frame(matrix(NA,nrow=length(ids),ncol=length(vars)))
  rownames(merged_data)=ids
  colnames(merged_data)=vars
  
  
  # Merge all datasets
  for (i in 1:length(ldata)) {
    merged_data[rownames(ldata[[i]]),colnames(ldata[[i]])] = ldata[[i]]

  }
  
  #merged_data$year <- year
  
  #rownames(merged_data) = merged_data[,"SEQN"]
  
  # Save to CSV
  write.csv(merged_data, sprintf("%s/nhanes_%s.csv", final_dir,name), row.names = FALSE)
  
}


# Function to download and parse variable names from documentation
download_and_extract_variable_names <- function(dest_dir,component="Laboratory",name="lab") 
{
  # URL of the NHANES 2001–2002 laboratory variable list
  #url <- sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/variablelist.aspx?Component=%s&Cycle=%s#print",component,year)
  url = sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/variablelist.aspx?Component=%s&Cycle=",component)
  
  # Read the HTML content from the URL
  page <- read_html(url)

  # Extract the table containing the variable list
  variable_table <- page %>%
    html_node("table") %>%  # Select the first table on the page
    html_table(fill = TRUE) # Convert the HTML table to a data frame

  # Display the first few rows of the extracted table
  #head(variable_table)

  variable_table = as.data.frame(variable_table)

  #drop duplicates
  variable_table = variable_table[!duplicated(variable_table[,1]),]
  rownames(variable_table) = variable_table[,1]

  write.csv(variable_table,sprintf("%s/nhanes_%s_vars.csv",dest_dir,name))
  return(variable_table)
}
```

```{r}
library(rvest)
library(dplyr)
library(stringr)

library(rvest)
library(dplyr)
library(stringr)

GetCodebookUrlsNew <- function(component = "Questionnaire") {
  # Get the webpage URL
  cycle_url <- sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=%s", component)
  
  # Read the webpage
  webpage <- read_html(cycle_url)
  
  # Extract the column names from the table header (<th>)
  column_names <- webpage %>%
    html_nodes("table th") %>%
    html_text(trim = TRUE)
  
  # Extract all rows (<tr>) from the table, skipping the first (junk) row
  rows <- webpage %>%
    html_nodes("table tr") %>%
    .[-1]  # Remove the first row
  
  # Process each row to extract text and links
  table_data <- lapply(rows, function(row) {
    # Extract row text (all cells)
    row_text <- row %>%
      html_nodes("td") %>%
      html_text(trim = TRUE)
    
    # Extract the first link in the row (if any)
    link <- row %>%
      html_node("a") %>%
      html_attr("href")
    
    # Return a combined list of row text and link
    c(row_text, url = ifelse(is.null(link), NA, paste0("https://wwwn.cdc.gov", link)))
    #c(row_text, url = ifelse(is.null(link), NA, paste0("https://wwwn.cdc.gov", sub("\\.htm$", "#Codebook.htm", link))))
  })
  
  # Convert the list into a data frame
  table_df <- do.call(rbind.data.frame, table_data)
  
  # Assign column names
  if (length(column_names) == ncol(table_df) - 1) {
    colnames(table_df) <- c(column_names, "url")
  } else {
    colnames(table_df) <- c(paste0("Column", seq_len(ncol(table_df) - 1)), "url")
  }
  
  # Return the final table with links
  return(table_df)
}
GetCodebookUrls <- function(component="Questionnaire") {
  #gets urls and some additional info
  cycle_url <- sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=%s",component)
  
  webpage <- read_html(cycle_url)
  
  table <- webpage %>%
    html_node("table") %>%
    html_table(fill = TRUE)

  # Extract the links (doc file URLs)
  links <- webpage %>%
    html_nodes("table a") %>%
    html_attr("href") %>%
    str_subset("(?i)\\.htm$") #?i makes it ignore case #,negate=FALSE


  # Create full URLs
  links <- paste0("https://wwwn.cdc.gov", links)
  
  #print(colnames(table))
  
  table = as.data.frame(table)
  table[,"url"] = links
  
  return(table)
}


GetCodebookUrlsOld <- function(component="Questionnaire") {
  #just gets urls
  cycle_url <- sprintf("https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=%s",component)
  
  #old
  
  # Scrape the page to find .XPT file links
  page <- read_html(cycle_url)
  links <- page %>%
    html_nodes("a") %>%
    html_attr("href") %>%
    str_subset("DataFiles") %>%
    str_subset("(?i)\\.htm$") #?i makes it ignore case
  
  #print(cycle_url)
  #print(links)
  if (length(links) == 0) {
      warning(sprintf("No .htm links found for component %s.", component))
  }
  
  # Create full URLs
  links <- paste0("https://wwwn.cdc.gov", links)
  return(links)
}


GetCodebook = function(url)
{
  #url <- "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2005/DataFiles/DEMO_D.htm#Codebook"
    #"https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/PFQ_E.htm#Codebook"
  #codebook has summary table including encoding, variable name and description

# Read the HTML content
page <- read_html(url)


#first make sure codebook is present
# Find all elements with ID "Codebook"
codebook_elements <- page %>% html_elements("#Codebook")

# Check if the "Codebook" section exists
if (length(codebook_elements) == 0) {
  warning(sprintf("Failed to find codebook for %s",url))
  return(list(summaries="failed",results="failed"))
}

# Find the "Codebook" section
codebook_div <- page %>% html_node("#Codebook")


# Extract each variable entry
entries <- codebook_div %>% html_nodes(".pagebreak")

# Initialize a list to store results
results <- list()

for (entry in entries) {
  #print(entry)
  # Extract the variable title
  vartitle <- entry %>%
    html_node(".vartitle") %>%
    html_text(trim = TRUE)
  
  # Extract key-value pairs
  dl_items <- entry %>%
    html_nodes("dl") %>%
    html_children()
  
  # Define specific fields to extract
  #keys <- dl_items %>%
  #  html_nodes("dt") %>%
  #  html_text(trim = TRUE)
  keys <- dl_items[seq(1, length(dl_items), by = 2)] %>%  # Extract every odd element (<dt>)
      html_text(trim = TRUE)
  
  #values <- dl_items %>%
  #  html_nodes("dd") %>%
  #  html_text(trim = TRUE)
  values <- dl_items[seq(2, length(dl_items), by = 2)] %>%  # Extract every even element (<dd>)
      html_text(trim = TRUE)
  
  #print(data.frame(k=keys,v=values))
  
  #check for repeated keys
  unkey = unique(keys)
  for (jj in 1:length(unkey))
  {
    logi = keys==unkey[jj]
    values[logi] = paste(values[logi],collapse=";")
  }
  values = values[!duplicated(keys)]
  keys = keys[!duplicated(keys)]
  
  #print(data.frame(k=keys,v=values))
  
  #print("make tibble")
  # Filter for specific keys
  info <- tibble(Key = keys, Value = values)
  
  # Pivot data for easier use
  info_wide <- info %>%
    pivot_wider(names_from = Key, values_from = Value)
  
  #print("k")
  #print(info_wide$Target)
  #print(str(info_wide))
  #print(info_wide)
  #print(do.call(c,info_wide$Target))  

  #make it a dataframe, easier for me to work with
  info_wide = as.data.frame(info_wide)
  if(!("Target:"%in%colnames(info_wide))) info_wide[,"Target:"] = NA
  
  # Extract the table (if present)
  table <- entry %>%
    html_node("table")
  
  if (!is.null(table)) {
    table_df <- tryCatch(
      {
        table %>% html_table(fill = TRUE)
      },
      error = function(e) {
        NULL
      }
    )
  } else {
    table_df <- NULL
  }


  #print(table_df)
  #print('make df')
  #make it a dataframe, easier for me to work with
  if(!is.null(table_df))
  {
    table_df = as.data.frame(table_df)
    table_df[,"Target"] = info_wide[1,"Target:"]
  }
  
  #print('summary df')
  #make a merged summary table
  summary_df = data.frame(name=as.character(info_wide[1,1]))
  summary_df[1,"nobs"] = NA
  summary_df[1,"nmissing"] = NA
  summary_df[1,"description"]=vartitle
  summary_df[1,"target"] = info_wide[1,"Target:"]
  summary_df[1,"codes"] = paste(table_df[,"Code or Value"],collapse=";;")
  summary_df[1,"values"] = paste(table_df[,"Value Description"],collapse=";;")
  summary_df[1,"count"] = paste(table_df[,"Count"],collapse=";;")
  summary_df[1,"cumulative"] = paste(table_df[,"Cumulative"],collapse=";;")
  temp = table_df[,"Skip to Item"]
  summary_df[1,"recode"] = NA
  summary_df[1,"skipto"] = paste(temp[!is.na(temp)],collapse=";;")
  summary_df[1,"gated"] = NA
  summary_df[1,"gated_by"] = NA
  summary_df[1,"datatype"] = NA
  #special cases:
  #if(setequal(c("yes","no","refused","don't know","missing"),tolower(table_df[,"Value Description"]))) 
  #print("special cases")
  if(!is.null(table_df))
  {
    if(nrow(table_df)==5)
    {
      if(all(c("yes","no","refused","don't know","missing")==tolower(table_df[,"Value Description"])))  #order matters
      {
        summary_df[1,"datatype"] = "binary"
        summary_df[1,"recode"] = paste(c(1,0,NA,NA,NA),collapse=";;")
        summary_df[1,"nobs"] = sum(table_df[1:2,"Count"])
        summary_df[1,"nmissing"] = sum(table_df[-(1:2),"Count"])
      } else if(all(c("yes","no","refuse","don't know","missing")==tolower(table_df[,"Value Description"])))  #order matters
      {
        summary_df[1,"datatype"] = "binary"
        summary_df[1,"recode"] = paste(c(1,0,NA,NA,NA),collapse=";;")
        summary_df[1,"nobs"] = sum(table_df[1:2,"Count"])
        summary_df[1,"nmissing"] = sum(table_df[-(1:2),"Count"])
      }
    } 
    else if(nrow(table_df)==8)
    {
      if(all(c("no difficulty","a little difficulty","much difficulty","unable to do","do not do this activity","refused","don't know","missing")==tolower(table_df[,"Value Description"])))  #order matters
      {
        summary_df[1,"datatype"] = "ordinal"
        summary_df[1,"recode"] = paste(c(0,1/3,2/3,1,NA,NA,NA,NA),collapse=";;")
        summary_df[1,"nobs"] = sum(table_df[1:4,"Count"])
        summary_df[1,"nmissing"] = sum(table_df[-(1:4),"Count"])
      }
      else if(all(c("no difficulty","some difficulty","much difficulty","unable to do","do not do this activity","refused","don't know","missing")==tolower(table_df[,"Value Description"])))
      {
        summary_df[1,"datatype"] = "ordinal"
        summary_df[1,"recode"] = paste(c(0,1/3,2/3,1,NA,NA,NA,NA),collapse=";;")
        summary_df[1,"nobs"] = sum(table_df[1:5,"Count"])
        summary_df[1,"nmissing"] = sum(table_df[-(1:5),"Count"])
      }
      else if(all(c("excellent","good","fair","poor","very poor?","refused","don't know","missing")==tolower(table_df[,"Value Description"])))
      {
        summary_df[1,"datatype"] = "ordinal"
        summary_df[1,"recode"] = paste(c(0,1/4,2/4,3/4,1,NA,NA,NA),collapse=";;")
        summary_df[1,"nobs"] = sum(table_df[1:5,"Count"])
        summary_df[1,"nmissing"] = sum(table_df[-(1:5),"Count"])
      }
    }
    else if(nrow(table_df)==2)
    {
      if(all(c("range of values","missing")==tolower(table_df[,"Value Description"])))  #order matters
      {
        summary_df[1,"datatype"] = "numeric"
        summary_df[1,"nobs"] = sum(table_df[1,"Count"])
        summary_df[1,"nmissing"] = sum(table_df[-1,"Count"])
      }
    }
    else if(nrow(table_df)==4)
    {
      if(all(c("range of values","refused","don't know","missing")==tolower(table_df[,"Value Description"])))  #order matters
      {
        summary_df[1,"datatype"] = "numeric"
        summary_df[1,"recode"] = paste(c("x",NA,NA,NA ),collapse=";;")
        summary_df[1,"nobs"] = sum(table_df[1,"Count"])
        summary_df[1,"nmissing"] = sum(table_df[-1,"Count"])
      }
    }
    #else if("range of values"%in%tolower(table_df[,"Value Description"])) # too general
    #{
    #  summary_df[1,"datatype"] = "numeric"
    #  summary_df[1,"nobs"] = sum(table_df[1,"Count"])
    #}
  }

  #print('done')

  # Store results in a list
  #print(info_wide)
  varname = as.character(info_wide[1,1])
  results[[varname]] <- list(
    name=varname,
    title=vartitle,
    info = info_wide,
    table = table_df,
    summary = summary_df
  )
}

#check for gating and merge summaries as you go
gatedNow = F
gateUntil = ""
gate = ""
summaries = list()
results[[1]]$summary[,"gated"] = F
for (i in 1:length(results))
{
  #print(results[[i]])
  if(gatedNow & gateUntil==results[[i]]$name)
  {
    gateNow=F
    gateUntil=""
    results[[i]]$summary[,"gated"] = F
  } else if (gatedNow)
  {
    results[[i]]$summary[,"gated"] = T
    results[[i]]$summary[,"gated_by"] = gate
  } else
  {
    results[[i]]$summary[,"gated"] = F
  }
  
  
  sk = results[[i]]$table[,"Skip to Item"]
  sk = sk[sk!=""]
  sk = sk[!is.na(sk)]
  
  if(length(sk) > 0)
  {
    gatedNow=T
    gateUntil = sk[1]
    gate = results[[i]]$name
  }
  
  summaries[[i]] = results[[i]]$summary
  summaries[[i]][,"order"] = i
}

summaries = do.call(rbind,summaries)


return(list(summaries=summaries,results=results))

# Combine all variable info into a single data frame
#all_info <- bind_rows(lapply(results, function(x) x$Info), .id = "Variable")

# Combine all tables into a single data frame
#all_tables <- bind_rows(
#  lapply(results, function(x) if (!is.null(x$Table)) x$Table else NULL),
#  .id = "Variable"
#)

# Save the results to CSV files
#write.csv(all_info, "NHANES_Codebook_Info.csv", row.names = FALSE)

#if (nrow(all_tables) > 0) {
#  write.csv(all_tables, "NHANES_Codebook_Tables.csv", row.names = FALSE)
#}

# Preview the extracted information
#print("Extracted Variable Information:")
#print(head(all_info))

#if (nrow(all_tables) > 0) {
#  print("Extracted Tables:")
#  print(head(all_tables))
#}

}
```

codebooks
```{r}
demourls=GetCodebookUrls("Demographics")

democodebook=list()
for (i in 1:nrow(demourls))
{
  democodebook[[i]] = GetCodebook(demourls[i,"url"])$summaries
  democodebook[[i]][,"years"] = demourls[i,"Years"]
  democodebook[[i]][,"class"] = demourls[i,"Data File Name"]
  democodebook[[i]][,"filename"] = demourls[i,"Doc File"]
  democodebook[[i]][,"url"] = demourls[i,"url"]
}
democodebook = do.call(rbind,democodebook)

save=T
if(save)
{
  write.csv(democodebook,sprintf("C:/data/nhanes/nhanes_demo/nhanes_demo_codebook.csv"))
}
```

```{r}
clinicurls=GetCodebookUrls("Questionnaire")

cliniccodebook=list()
ind = 1
for (i in 1:nrow(clinicurls))
{
  temp = GetCodebook(clinicurls[i,"url"])$summaries
  if(all(temp=="failed")) next
  cliniccodebook[[ind]] = temp
  cliniccodebook[[ind]][,"years"]    = clinicurls[i,"Years"]
  cliniccodebook[[ind]][,"class"]    = clinicurls[i,"Data File Name"]
  cliniccodebook[[ind]][,"filename"] = clinicurls[i,"Doc File"]
  cliniccodebook[[ind]][,"url"]      = clinicurls[i,"url"]
  ind = ind + 1
}
cliniccodebook = do.call(rbind,cliniccodebook)

save=T
if(save)
{
  write.csv(cliniccodebook,sprintf("C:/data/nhanes/nhanes_clinic/nhanes_clinic_codebook.csv"))
}
```

```{r}
laburls=GetCodebookUrlsNew("Laboratory")

labcodebook=list()
ind = 1
for (i in 1:nrow(laburls))
{
  temp = GetCodebook(laburls[i,"url"])$summaries
  if(all(temp=="failed")) next
  labcodebook[[ind]] = temp
  labcodebook[[ind]][,"years"]    = laburls[i,"Years"]
  labcodebook[[ind]][,"class"]    = laburls[i,"Data File Name"]
  labcodebook[[ind]][,"filename"] = laburls[i,"Doc File"]
  labcodebook[[ind]][,"url"]      = laburls[i,"url"]
  ind = ind + 1
}
labcodebook = do.call(rbind,labcodebook)

save=T
if(save)
{
  write.csv(labcodebook,sprintf("C:/data/nhanes/nhanes_lab/nhanes_lab_codebook.csv"))
}
```

update: get dietary data (May 2025)
```{r}
dieturls=GetCodebookUrlsNew("Dietary")

dietcodebook=list()
ind = 1
for (i in 1:nrow(dieturls))
{
  temp = GetCodebook(dieturls[i,"url"])$summaries
  if(all(temp=="failed")) next
  dietcodebook[[ind]] = temp
  dietcodebook[[ind]][,"years"]    = dieturls[i,"Years"]
  dietcodebook[[ind]][,"class"]    = dieturls[i,"Data File Name"]
  dietcodebook[[ind]][,"filename"] = dieturls[i,"Doc File"]
  dietcodebook[[ind]][,"url"]      = dieturls[i,"url"]
  ind = ind + 1
}
dietcodebook = do.call(rbind,dietcodebook)

save=T
if(save)
{
  write.csv(dietcodebook,sprintf("C:/data/nhanes/nhanes_diet/nhanes_diet_codebook.csv"))
}
```

demographics
```{r}
  download_and_save_files(dest_dir="C:/data/nhanes/nhanes_demo/temp",
                            final_dir="C:/data/nhanes/nhanes_demo",name="demo",component="Demographics")

  download_and_extract_variable_names(dest_dir="C:/data/nhanes/nhanes_demo",component="Demographics",name="demo")
```

lab
```{r}
  download_and_save_files(dest_dir="C:/data/nhanes/nhanes_lab/temp",
                            final_dir="C:/data/nhanes/nhanes_lab",name="lab",component="Laboratory")

  download_and_extract_variable_names(dest_dir="C:/data/nhanes/nhanes_lab",component="Laboratory",name="lab")
```

questionnaire
```{r}
  download_and_save_files(dest_dir="C:/data/nhanes/nhanes_clinic/temp",
                            final_dir="C:/data/nhanes/nhanes_clinic",name="clinic",component="Questionnaire")


  download_and_extract_variable_names(dest_dir="C:/data/nhanes/nhanes_clinic",component="Questionnaire",name="clinic")

```

performance / exam
```{r}
  download_and_save_files(dest_dir="C:/data/nhanes/nhanes_exam/temp",
                            final_dir="C:/data/nhanes/nhanes_exam",name="exam",component="Examination")


  download_and_extract_variable_names(dest_dir="C:/data/nhanes/nhanes_exam",component="Examination",name="exam")

```

LimitedAccess 
```{r}
#  download_and_save_files(dest_dir="C:/data/nhanes/nhanes_misc/temp",
#                            final_dir="C:/data/nhanes/nhanes_misc",name="misc",component="LimitedAccess")


#  download_and_extract_variable_names(dest_dir="C:/data/nhanes/nhanes_misc",component="LimitedAccess",name="misc")

```

diet - added May 2025
```{r}
  download_and_save_files(dest_dir="C:/data/nhanes/nhanes_diet/temp",
                            final_dir="C:/data/nhanes/nhanes_diet",name="diet",component="Dietary")


  download_and_extract_variable_names(dest_dir="C:/data/nhanes/nhanes_diet",component="Dietary",name="diet")

```

# old? (not sure???)
# code books

```{r}
examurls=GetCodebookUrlsNew("Examination")

examcodebook=list()
ind = 1
for (i in 1:nrow(examurls))
{
  if(grepl("withdrawn",tolower(examurls[i,"Date Published"])))
  {
    warning(sprintf("skipping %s, withdrawn",examurls[i,"Date File Name"]))
    next
  }
  temp = GetCodebook(examurls[i,"url"])$summaries
  if(all(temp=="failed")) next
  examcodebook[[ind]] = temp
  examcodebook[[ind]][,"years"]    = examurls[i,"Years"]
  examcodebook[[ind]][,"class"]    = examurls[i,"Data File Name"]
  examcodebook[[ind]][,"filename"] = examurls[i,"Doc File"]
  examcodebook[[ind]][,"url"]      = examurls[i,"url"]
  ind = ind + 1
}
examcodebook = do.call(rbind,examcodebook)

save=T
if(save)
{
  write.csv(examcodebook,sprintf("C:/data/nhanes/nhanes_exam/nhanes_exam_codebook.csv"))
}
```

#old code
```{r}
stop("old code")
```

to add additional info to a tibble
sample code
```{r}
library(tibble)

# Example data: Variable values for individuals
data <- tibble(
  PFQ054 = c(1, 2, 1, 9),  # Example responses for "PFQ054"
  PFQ055 = c(2, 2, 1, 1)   # Example responses for "PFQ055"
)

# Add metadata as attributes for each column
attr(data$PFQ054, "Variable Name") <- "PFQ054"
attr(data$PFQ054, "SAS Label") <- "Need special equipment to walk"
attr(data$PFQ054, "English Text") <- "Because of a health problem, do you have difficulty walking without using any special equipment?"
attr(data$PFQ054, "Target") <- "Both males and females 20 YEARS - 150 YEARS"

attr(data$PFQ055, "Variable Name") <- "PFQ055"
attr(data$PFQ055, "SAS Label") <- "Need special help to climb stairs"
attr(data$PFQ055, "English Text") <- "Do you have difficulty climbing stairs?"
attr(data$PFQ055, "Target") <- "Both males and females 30 YEARS - 150 YEARS"

# View the main data
print(data)

# Access metadata for PFQ054
print(attributes(data$PFQ054))

```

scraping code book
example
```{r}
library(rvest)
 
 # URL of the codebook
 url <- "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2005/DataFiles/PFQ_D.htm"
 
 # Scrape the page
 page <- read_html(url)
 
 # Extract table content (variables, descriptions, and value labels)
 variables <- page %>%
     html_nodes("table") %>%  # Find all tables
     .[[1]] %>%              # Assume the first table is what you need
     html_table(fill = TRUE) # Convert to a data frame
 
 # Preview the variables
 print(variables)
```




no longer relevant
```{r}
nhanes_years = c(
  "1999-2000", "2001-2002", "2003-2004", "2005-2006", 
  "2007-2008", "2009-2010", "2011-2012", "2013-2014", 
  "2015-2016", "2017-2018" #, "2019-2020" #, "2021-2023" #not on RNHANES
)
```

download demographics
```{r}
for (i in 1:length(nhanes_years))
{
  message("Downloading year: ", nhanes_years[i])
  download_and_save_files(nhanes_years[i],dest_dir="C:/data/nhanes/nhanes_demo/temp",
                            final_dir="C:/data/nhanes/nhanes_demo",name="demo",component="Demographics")
}
```

download demographics documentation
```{r}
for (i in 1:length(nhanes_years))
{
  message("Downloading year: ", nhanes_years[i])
  download_and_extract_variable_names(nhanes_years[i],dest_dir="C:/data/nhanes/nhanes_demo",component="Demographics",name="demo")
}
```

download lab
```{r}
for (i in 1:length(nhanes_years))
{
  message("Downloading year: ", nhanes_years[i])
  download_and_save_files(nhanes_years[i],dest_dir="C:/data/nhanes/nhanes_lab/temp",
                            final_dir="C:/data/nhanes/nhanes_lab",name="lab",component="Laboratory")
}
```

download lab documentation
```{r}
for (i in 1:length(nhanes_years))
{
  message("Downloading year: ", nhanes_years[i])
  download_and_extract_variable_names(nhanes_years[i],dest_dir="C:/data/nhanes/nhanes_lab",component="Laboratory",name="lab")
}
```

#questionaire data
```{r}
for (i in 1:length(nhanes_years))
{
  message("Downloading year: ", nhanes_years[i])
  download_and_save_files(nhanes_years[i],dest_dir="C:/data/nhanes/nhanes_clinic/temp",
                            final_dir="C:/data/nhanes/nhanes_clinic",name="clinic",component="Questionnaire")
}
```

download questionnaire documentation
```{r}
for (i in 1:length(nhanes_years))
{
  message("Downloading year: ", nhanes_years[i])
  download_and_extract_variable_names(nhanes_years[i],dest_dir="C:/data/nhanes/nhanes_clinic",component="Questionnaire",name="clinic")
}
```
